### 2025-06-28 paper/sections/method.qmd
Adjust methods

In this section we propose our novel counterfactual training objective. In CT, we combine ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of aligning representations with plausible explanations that comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. To be actionable, AR recommendations must also be attainable. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

@@

\begin{enumerate}
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
\end{enumerate}
where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable}. 
\end{definition}

`\noindent`{=latex} The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

To train models with high explainability as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$. The first term can be optimized using conventional cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your relies on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology. Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

The departure from SGLD of [@grathwohl2020your] allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods facilitate the imposition of domain and mutability constraints. In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility wrt. the choice for $\text{advloss}(\cdot)$ in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing that defines an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing that considers misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

@@

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.

Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. We have $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
3. We have $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

`\noindent`{=latex} It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.

We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT.

@@

In all cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green)---but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to their ground-truth label: $y^-=1$ in blue and $y^+=2$ in orange. Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

>>>>

This section introduces the Counterfactual Training objective. CT combines ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of producing models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.

In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which generated CEs comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face]. Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models. An AR recommendations must also be actionable, i.e., possible to attain by the recipient. A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they had been 20 years older. Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults. But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature. Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability. Formally, we define explainability as follows:

\begin{definition}[Model Explainability]
\label{def-explainability}
Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$, where $\arg\max_y{\mathbf{y}^{+}}=y^+$ is the index of the target class.

@@

\begin{enumerate}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$, where $A$ is some arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
\end{enumerate}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

`\noindent`{=latex} The characterization of faithfulness and plausibility in Def. \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation. Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data. Actionability constraints in Def. \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed. In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$. We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.

@@

To train explainable models as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:

$$
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
$$ {#eq-obj}
where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy). The second and third terms are explained in detail below. For now, they can be summarized as inducing explainability directly and indirectly by penalizing (in 2\textsuperscript{nd} term) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (in 3\textsuperscript{rd} term) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$. Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit]. The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.

@@

@grathwohl2020your observed that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$. @grathwohl2020your show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj. To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.

A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def. \ref{def-explainability}. This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$. To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology. Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples. Specifically, we have:

@@

The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata. For example, many methods support domain and mutability constraints. In principle, any existing approach for generating CEs is viable, so long as it does not violate the faithfulness condition. Like JEMs [@murphy2022probabilistic], Counterfactual Training can be considered a form of contrastive representation learning.

@@

Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring]. This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj. An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:

$$
\begin{aligned}
\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
\end{aligned}
$$ {#eq-adv}
Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$. This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation". Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE. One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.

@@

Many existing counterfactual explainers support domain and mutability constraints out-of-the-box. In fact, both types of constraints can be implemented for any explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining]. In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain. Similarly, mutability constraints can be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations. Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div. Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints. For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints. Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:

1. Feature $d$ is strictly immutable in practice.
2. $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but $d$ can only be decreased in practice.
3. $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but $d$ can only be increased in practice.

`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$. Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features. This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments. Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]

\begin{proposition}[Protecting Immutable Features]
\label{prp-mtblty}
Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$. Assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
\end{proposition}

`\noindent`{=latex} It is worth noting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features. This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected. We address this in @sec-discussion.

@@

Suppose we are interested in predicting the likelihood that loan applicants default on their credit. We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable. We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.

@@

In all four cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green) into the target class---but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange). Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
